\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-01-23 document/XSGRB_sample.tex
Addressing Kaspers comments and notes on 2014 bursts

\abstract{In this work we present the spectra of all $\gamma$-ray bursts (GRBs)
	afterglows that have been promptly observed at the VLT/X-shooter up to
	\termdate. This totals spectroscopic observations of 102 individual bursts
	observed with 48 hours of the GRB trigger. Redshifts have been measured for 96
	\% of these, covering a redshift range from 0.059 to 7.84. Based on an set of
	selection criteria that are unbiased with regards to intrinsic properties of the
	GRBs, the follow-up effort has been guided by the \textit{Swift}-telescope to
	produce a homogeneous sample of afterglows for 86 spectra. We here provide a
	public release of all the reduced spectra, including continuum estimates and
	telluric corrections. For completeness, we also provide reductions for the
	late-time observations of the underlying host galaxies. We provide an assessment
	of the degree of completeness with respect to the parent GRB population, in
	terms of the X-ray properties of the bursts in the sample and find that the
	sample presented here is consistent with being drawn from the \textit{Swift}
	sample. For all bursts (39) for which it is possible, we provide a measurement
	of the hydrogen column density, increasing the total number of published HI
	column density measurements by $\sim$ 50 per cent. We constrain the fraction of
	dark bursts to be less than 28 per cent based on the detection of an optical
	afterglow and tentatively confirm previous results that larger optical darkness
	is correlated with increased X-ray absorption. This dataset provides a unique
	possibility to study the ISM across cosmic time. }

@@

Gamma-Ray bursts (GRBs) are interesting as they provide the opportunity to
derive new constraints on a very wide range of topics in astrophysics. Examples
range from small scale phenomena relating to magnetars, properties of highly
relativistic jets, hyper/supernova explosions, the interstellar medium, dust
extinction curves, starburst galaxies, chemical and molecular abundances, escape
of ionizing radiation, the ionization state of the intergalactic medium,
intervening absoption systems to standard candles in cosmology
\citep[e.g.,][]{Lyons2010, Molinari2007, HjorthBloom2012, Wijers1998,
	Prochaska2009, Savaglio2006, Ghirlanda2007}. In that sense GRB research is very
multidisciplinary.

The {\it Swift} satellite \citep{Gehrels2009}, which was launched in 2004, has
made it possible to harvest much of the great potential in using GRBs as probes
of galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.,][]{Ricker2004}. A crucial aspect of the great
success of the {\it Swift} mission has been the extensive ground based
follow-up observations of the afterglows and of the host galaxies of the GRBs.
This work has involved a large community of researchers both inside and outside
the official {\it Swift} team. This fruitful collaboration has been greatly
facilitated by the open data access policy of the {\it Swift} mission.

@@

Because the localization accuracy of BAT is 1 - 4 arcminutes
\citep{Barthelmy2005}, a secure host identification is in general impossible
based on BAT alone. We therefore additionally require an X-ray position from the
X-ray Telescope (XRT) \citep{Burrows2005} to be distributed to the GCN network
\citep{Barthelmy2000} within 10 hours and to account for observing constraints
on \textit{Swift} that XRT began observations within 10 minutes. The additional
timing requirement of the XRT follow-up means that all bursts in our sample have
detected X-ray afterglows. Because the XRT completeness is very high
\citep{Burrows2007}, this cut should not alter the parent sample significantly.

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to completed. Since the GRB population is isotropically distributed
on the sky as seen from Earth and because the GRB properties do not depend on
position on the sky \citep{Briggs1996}, this cut does not influence our ability
to fairly sample the underlying GRB population. The same arguments apply to
the requirement that there are no nearby bright stars. We additionally require
that the Galactic extinction is below $A_V \lesssim 0.5$ mag, based on the
extinction maps by \citet{Schlegel1998}. Because of isotropy, these additional
cuts should not influence the optical properties of the bursts themselves, only
our ability to successfully secure observations that allows us to investigate
the spectroscopic properties of GRBs.

@@

The use of RRM is unique in the sense that it allowes for the detailed study of
the temporal variability of the GRB afterglow due to effect of the GRB itself on
the sorrounding medium \citep[e.g., see][for examples]{Dessauges-Zavadsky2006,
	Vreeswijk2007, Vreeswijk2013}

@@

The first GRB followed up is GRB~090313, observed on the 15th of March, 2009,
during the commissioning of X-shooter UT3. The bursts observed during the
commissioning or science verification (GRB~090313, GRB~090530, GRB~090809,
GRB~090926) are not a part of our statistical sample. The first burst observed
after science verification and X-shooter was moved to UT2, is GRB~091018, which
thereby constitute the first burst entering our statistical sample. For all
bursts that fulfill our sample selection criteria, described in Sect.
\ref{samplecrit}, spectroscopic follow-up has been attempted with X-shooter.
Various conditions can affect our ability to follow up a given burst, and a
discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

@@

For the majority of the bursts, we have observed with a slit width of 1\farc0,
0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively. This sets a
lower limit on the delivered resolution of the spectra based on the tabulated
values of the delivered resolutions, which is 4350, 7450, and 5300 for the UVB,
VIS and NIR-arm
respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}. %The slit width sets the width of the atmospheric sky lines and determines the amount of light lost due to the wavelength-dependent seeing PSF extending outside the coverage of the slit, where the width of sky-lines is always set by the slit width whereas both the delivered resolution and the slitloss changes for the better as the seeing PSF drops below the slit width. For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a 0\farc9 slit only 76.1 percent of the light will enter the slit, meaning that for almost all observations a slitloss correction is required.
%We describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements involving line profiles, knowledge of the precise
instrumental resolution is required. The resolution becomes better than the
nominal one, listed above, when the delivered seeing is smaller than the
projected width of the slit on the sky. We discuss how we determine the
effective instrumental resolution in Sect. \ref{resolution}.

@@

We provide an overview of all the observations in Table
\ref{tab:sample_overview} and plot the positions of all the bursts on the
celestial sphere in galactic coordinates in Fig. \ref{fig:skymap}. Away from
central zone of avoidance due to high galactic extinction cutoff (marked in
white), the GRB positions have an apparent isotropic distribution except in the
upper left quadrant, which cannot be observed from Paranal.

@@

The basis for the reductions is the VLT/X-shooter pipeline, version
\texttt{2.7.1} or newer \citep{Goldoni2006, Modigliani2010}. The pipeline is
managed with the Reflex interface \citep{Freudling2013} and is used for
subtraction of bias level, flat-fielding, tracing of the echelle orders,
wavelength calibrations with the use of arc-line lamps, flux calibration using
spectrophotometric standards \citep{Vernet2010, Hamuy1994}, mirror flexure
compensation, initial sky-subtraction and lastly the rectification and merging
of the orders. For the initial sky-subtraction, the background has been
estimated in regions adjacent to the object trace clear of contaminating
sources. Because of the broken ADC, for some objects there is a lot of curvature
in the object trace along the dispersion axis of the slit. This means that for
some bursts, the initial sky-estimate has been made from a limited number of
pixels in the spatial direction. By doing an initial subtracting the sky on the
un-rectified image we ensure that bulk of the sky background is not
redistributed by the rectification process.

The image is rectified onto an equidistant grid with a dispersion sampling of
0.02 nm/pixel and a 0.16 "/pixel spatial sampling for the UVB and VIS arm and
0.06 nm/pixel with a 0.21 "/pixel in the NIR arm.  Because the tabulated
resolution is a lower limit, by choosing a sampling of 0.02 nm/pixel, we ensure
that the bluest part of neither of the arms have a sampling lower than the
Nyquist sampling rate of 2 pixels per resolution FWHM.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as a single observation and then subsequently combined to form a
single image. Because this strategy is employed, we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images where the noise in each pixel is dominated by
Poisson noise it is important to estimate the background variance in a large enough
region, so that any correlation between the signal and the weights are removed.
To this end, the weight map is generated by a running median window over the
variance map, where the trace has been masked and width of the window is chosen
to be wide enough for median variance to be generated on the basis of several
hundred pixels. This weighting scheme automatically also optimally combines
images of different exposure times or images where the background is varying,
which is often the case when a burst has been observed close to twilight

@@

X-shooter, being installed at the VLT Cassegrain focus is  prone to flexures
during operations. The flexures modify the projection of the slit on the
detector with respect to the one obtained in daytime calibration. This requires
a modification of the wavelength solution in order to process correctly the
night-time data. Part of this correction is performed by the pipeline using the
frames taken during X-shooter Active Flexure Compensation procedure
\footnote{X-shooter User Manual available at
	https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}. The
remaining offset is corrected by cross-correlating the observed sky spectrum
with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum,
estimated as the mode of all flux values, has been subtracted. To get the
correct seeing PSF with which to convolve the synthetic sky an initial
refinement of the wavelength solution have been obtained by cross-correlating
the observed sky with an unconvolved synthetic sky. This preliminary wavelength
calibration is applied to the observed sky. The synthetic spectrum is then
convolved with an increasing seeing PSF and the width that minimizes $\chi^2$
with the updated observed sky is chosen to be the effective sky-PSF. Using the
synthetic sky with the matched resolution, a final wavelength calibration can
then be calculated by cross-correlating the observed sky with the correctly
broadened sky spectrum, as a function of a velocity offset. Both a
multiplicative and an additive offset to the wavelength calibration has been
tested, but in terms of $\chi^2$, the model with only a multiplicative offset is
preferred. The resulting offsets, which were smaller than 0.01 nm in the UVB and
VIS data and smaller than 0.05 nm in the NIR spectra, but changing over short
period of time were applied to the corresponding spectra. Using the convolved
synthetic sky, the pixels containing the brightest sky lines have been added to
the bad pixel map.

@@

To extract the afterglow spectrum from the rectified 2D-image, several
techniques have been employed based on the brightness of the afterglow and the
complexity of the objects entering the slit. Due to the malfunctioning ADC, see
Sect. \ref{obs}, the spectral trace changes position across the slit in the
spatial direction as a function of wavelength. For a large fraction of the
bursts observed, using a single aperture for the spectral extraction is
inadequate due to the large amount of background that would then enter the slit.
To optimally select the extraction regions we therefore need to model the trace
position.

To get the shape and the position of the spectral PSF as a function of location
on the image, we need to chose a model which can represent how the light falls
on the slit. We know from \citet{Trujillo2001} that the Moffat function
\citep{Moffat1969} adequately describes an imaging PSF due to atmospheric
turbulence, but because of aberrations in the dispersion elements and the
rectification process, the PSF we are trying to model is different from this
profile. To allow for flexibility in the model, we have chosen the Voigt
function as a model for the spectral PSF and we describe how this is evaluated
in App. \ref{voigt}. Since additionally, the host galaxy could also have a
contribution the image profile, this choice allows for the required freedom if
additional flux is in the wings of the profile.

@@

Based on the signal-to-noise of the afterglow continuum, the 2D-image has been
binned down in the spectral direction to a number of elements that allows for an
accurate tracing of the PSF, typically 200 bins for moderate signal-to-noise.
For each of the bins, using the analytically guided guess position, the spectral
PSF has been fit using the unweighted chi-squared minimization algorithm
implemented in \texttt{scipy.optimize.curve\_fit} \citep{scipy}. Since we know
that the trace varies slowly as a function of wavelength, we have then fitted a
low-order polynomial to the fit parameters as a function of wavelength, which
allows us to evaluate the spectral PSF at all wavelengths and this way
accurately model the entire spectral PSF.

Equipped with a model for how the light is distributed across the entire
dispersion direction, we can  employ the optimal extraction algorithm
\cite{Horne1986}, which weights the extraction aperture by the spectral profile,
or alternatively sum all pixels within 1 FWHM of the modeled profile. Where
possible, we have used the optimal extraction. In cases where the trace is very
weak, even in the binned images, an aperture has been selected manually which
covers emission lines, if present, and when nothing is immediately visible, the
entire nodding window. The error- and bad pixel maps are in all cases propagated
throughout the extraction.

@@

The spectra are corrected for Galactic extinction using the E(B-V) value from
the dust maps of \citet{Schlegel1998} with the update in
\citet{Schlafly2011}\footnote{Queried from
	http://irsa.ipac.caltech.edu/applications/DUST/index.html using
	\citet{astroquery}}, and the extinction curve by \cite{Cardelli1989} with a
total to selective extinction $R_V = 3.1$. The wavelengths of the extracted
1D-spectra are moved to vacuum, corrected for barycentric motion, and the
wavelength recalibration described in Sect. \ref{wavecal} is applied. Pixels
with pixel-to-pixel variation larger than $50 \sigma$ are additionally added to
the bad pixel map.

@@

For all earth-based telescopes, the light first has to pass through Earths
atmosphere, where the atmospheric content and conditions make an imprint on the
received spectrum. These telluric features can be corrected for in a multitude
of ways. We employ a prioritized list of methods here, depending on the
availability of the different method. Since the observation are often taken at
odd times under varying conditions, this prioritized list ensures that we are
always doing the optimal correction.

@@

In the last case, where the object is neither bright enough, or there for some
reason have not been observed a telluric standard, we rely on the synthetic sky
model by \citep{Noll2012, Jones2013} for which we generate a synthetic
transmission spectrum,  where the ambient parameters for the
observations have been used.

@@

The afterglow spectra described in this paper are obtained in
Target-of-Opportunity (override) mode. In most cases there is therefore little
possibility to tweak slit widths to the seeing at the time of observations (i.e.
to optimize spectral resolution and signal to noise), and almost all our data is
therefore taken with a fixed set of slit widths and binning, described above. In
a fair number of cases, the seeing full width at half maximum (FWHM) is
considerably smaller than the silt width, and the delivered spectral resolution
will then be determined by the seeing rather than slit width, as afterglows are
point sources (this is evidently not the case for extended sources, e.g for host
galaxies). The delivered resolution for slit-width-dominated spectra
post-reduction and extraction can easily be determined from the bright sky
emission lines. For afterglow spectra with very high signal to noise, the
delivered spectral resolution can at times be determined from the science data
themselves. However, in the presence of multiple velocity components in
absorption, other forms of line broadening, and a lack of lines at some
redshifts, this is difficult to do at poorer signal to noise ratios (the
majority of spectra in our sample). A broad starting value for the expect
resolution will help fitting of these spectra, and can be important in upper
limit determination, and for this reason we construct a crude relation between
the seeing and the delivered resolution at our slit width, binning, and
reduction pipeline settings.

@@

All the spectra are made available at
\url{http://www.dark-cosmology.dk/~jselsing/XSGRB} and additionally through the
ESO archive in the form of phase 3 material
\footnote{\url{http://archive.eso.org/wdb/wdb/adp/phase3_main/form}}. This
release includes the spectra of GRBs outside the statistical sample and includes
both prompt afterglow observations as well as late time observations of the
associated hosts and represents \textit{all} afterglow spectra of GRBs carried
out by VLT/X-shooter from the commissioning of the instrument, \startdate, and
until the end of the last period of the program 098.A-0055, \termdate~and thus
constitutes 8 years of GRB afterglow observations with VLT/X-shooter. An
overview of all the spectra provided here are given in Table.
\ref{tab:sample_overview}. For each burst, each individual observation is provided
in a separate reduction and in cases where observations have been repeated for
an increased S/N or to follow the temporal evolution, a combined spectrum is
also provided. No attempt has been made to join the spectroscopic arms, so for
each observation, three spectra are provided in separate files.

@@

Each file contains 7 columns with following contents and descriptions:
\begin{itemize}
	\item WAVE - Observed wavelength in vacuum, corrected for barycentric motion and drifts in the wavelength solution ($\mathrm{\AA}$).
	\item FLUX - Observed flux density ($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-1} \mathrm{\AA}^{-1}$)
	\item ERR - Associated flux density error ($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-1} \mathrm{\AA}^{-1}$)
	\item QUAL - Bad pixel map, where a value different from zero indicates a bad pixel.
	\item CONTINUUM - Continuum estimate based on Sect. \ref{continuum} ($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-1} \mathrm{\AA}^{-1}$)
	\item CONTINUUM\_ERR - Relative error on the continuum estimate 
	\item TELL\_CORR - Inverse transmission spectrum. Multiply FLUX and ERR column with this column to correct for telluric absorption. 
\end{itemize}

@@

In this section, we describe the successfulness of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
obtained sample is representative for the full \textit{Swift} sample. An
important note is that here we include \textit{all} GRB after \startdate, that
have been observed with X-shooter, while only a subset of these constitute our
\textit{statistical} sample. The statistical sample is based on the selection
criteria described in Sect. \ref{samplecrit}. Some bursts not fulfilling the
sample criteria have been followed up due to interesting characteristics, e.g.
curiosity properties of their light curves, their brightness, etc. These bursts
are not discussed as part of the investigation of the statistical properties of
the GRB population. A prime example of a burst outside the statistical sample is
bright INTEGRAL burst GRB~161023A, that turned to be a rich spectrum with at
least 11 intervening absorption systems.

@@

>>>>

\abstract{In this work we present the spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed at the VLT/X-shooter up to
	\termdate. This totals spectroscopic observations of 102 individual bursts
	observed with 48 hours of the GRB trigger. Redshifts have been measured for 96
	\% of these, covering a redshift range from 0.059 to 7.84. Based on a set of observational selection criteria that are unbiased with regards to intrinsic properties of the
	GRBs, the follow-up effort has been guided by the \textit{Swift}-telescope to
	produce a homogeneous sample of afterglows for 86 spectra. We here provide a
	public release of all the reduced spectra, including continuum estimates and
	telluric corrections. For completeness, we also provide reductions for the
	late-time observations of the underlying host galaxies. We provide an assessment
	of the degree of completeness with respect to the parent GRB population, in
	terms of the X-ray properties of the bursts in the sample and find that the
	sample presented here is consistent with being drawn from the \textit{Swift}
	sample. For all bursts (39) for which it is possible, we provide a measurement
	of the neutral hydrogen column density, increasing the total number of published HI
	column density measurements by $\sim$ 50 per cent. We constrain the fraction of
	dark bursts to be less than 28 per cent based on the detection of an optical
	afterglow and tentatively confirm previous results that larger optical darkness
	is correlated with increased X-ray absorption. This dataset provides a unique
	possibility to study the ISM across cosmic time. }

@@

$\gamma$-ray bursts (GRBs) are interesting as they provide the opportunity to
derive new constraints on a very wide range of topics in astrophysics. Examples
range from small scale phenomena relating to magnetars, properties of highly
relativistic jets, hyper/supernova explosions, the interstellar medium, dust
extinction curves, starburst galaxies, chemical and molecular abundances, escape
of ionizing radiation, the ionization state of the intergalactic medium,
intervening absoption systems to standard candles in cosmology
\citep[e.g.,][]{Lyons2010, Molinari2007, HjorthBloom2012, Wijers1998,
	Prochaska2009, Savaglio2006, Ghirlanda2007}. In that sense GRB research is very
multidisciplinary.

The \textit{Niel Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by results
from earlier missions \citep[e.g.,][]{Ricker2004}. A crucial aspect of the great
success of the {\it Swift} mission has been the extensive ground based follow-up
observations of the afterglows and of the host galaxies of the GRBs. This work
has involved a large community of researchers both inside and outside the
official {\it Swift} team. This fruitful collaboration has been greatly
facilitated by the open data access policy of the {\it Swift} mission.

@@

Because the localization accuracy of BAT is 1 - 4 arcminutes
\citep{Barthelmy2005}, a secure host identification is in general impossible
based on BAT alone. We therefore additionally require an X-ray position from the
X-Ray Telescope (XRT) \citep{Burrows2005} to be distributed to the GCN network
\citep{Barthelmy2000} within 12 hours and to account for observing constraints
on \textit{Swift} that XRT began observations within 10 minutes. The additional
timing requirement of the XRT follow-up means that all bursts in our sample have
detected X-ray afterglows. Because the XRT completeness is very high
\citep{Burrows2007}, this cut should not alter the parent sample significantly.

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to be completed. Since the GRB population is isotropically distributed
on the sky as seen from Earth and because the GRB properties do not depend on
position on the sky \citep{Meegan1992, Briggs1996}, this cut does not influence our ability
to fairly sample the underlying GRB population. The same arguments apply to
the requirement that there are no nearby bright stars. We additionally require
that the Galactic extinction is below $A_V \lesssim 0.5$ mag, based on the
extinction maps by \citet{Schlegel1998}. Because of isotropy, these additional
cuts should not influence the optical properties of the bursts themselves, only
our ability to successfully secure observations that allows us to investigate
the spectroscopic properties of GRBs.

@@

The use of RRM is unique in the sense that it allows for the detailed study of
the temporal variability of the GRB afterglow due to effect of the GRB itself on
the sorrounding medium \citep[e.g., see][for examples]{Dessauges-Zavadsky2006,
	Vreeswijk2007, Vreeswijk2013}

@@

The first GRB followed up is GRB~090313, observed on the 15th of March, 2009,
during the commissioning of X-shooter on UT3. The bursts observed during the
commissioning or science verification (GRB~090313, GRB~090530, GRB~090809,
GRB~090926) are not a part of our statistical sample. The first burst observed
after science verification and X-shooter was moved to UT2, is GRB~091018, which
thereby constitute the first burst entering our statistical sample. For all
bursts that fulfill our sample selection criteria, described in Sect.
\ref{samplecrit}, spectroscopic follow-up has been attempted with X-shooter.
Various conditions can affect our ability to follow up a given burst, and a
discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

@@

For the majority of the bursts, we have observed with a slit width of 1\farc0,
0\farc9, and 0\farc9 for the UVB, VIS, and NIR-arm respectively. This sets a
lower limit on the delivered resolution of the spectra based on the tabulated
values of the delivered resolutions, which is 4350, 7450, and 5300 for the UVB,
VIS and NIR-arm
respectively\footnote{\url{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}}. %The slit width sets the width of the atmospheric sky lines and determines the amount of light lost due to the wavelength-dependent seeing PSF extending outside the coverage of the slit, where the width of sky-lines is always set by the slit width whereas both the delivered resolution and the slitloss changes for the better as the seeing PSF drops below the slit width. For atmospheric conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a 0\farc9 slit only 76.1 per cent of the light will enter the slit, meaning that for almost all observations a slitloss correction is required.
%We describe how slitlosses were corrected for in Sect. \ref{postproc}.
For accurate measurements involving line profiles, knowledge of the precise
instrumental resolution is required. The resolution becomes better than the
nominal one, listed above, when the delivered seeing is smaller than the
projected width of the slit on the sky. We discuss how we determine the
effective instrumental resolution in Sect. \ref{resolution}.

@@

We provide an overview of all the observations in Table
\ref{tab:sample_overview} and plot the positions of all the bursts on the
celestial sphere in galactic coordinates in Fig. \ref{fig:skymap}. Away from
central zone of avoidance due to high Galactic extinction cutoff (marked in
white), the GRB positions have an apparent isotropic distribution except in the
upper left quadrant, which cannot be observed from Paranal.

@@

The basis for the reductions is the VLT/X-shooter pipeline, version
\texttt{2.7.1} or newer \citep{Goldoni2006, Modigliani2010}. The pipeline is
managed with the Reflex interface \citep{Freudling2013} and is used for
subtraction of bias level, flat-fielding, tracing of the echelle orders,
wavelength calibrations with the use of arc-line lamps, flux calibration using
spectrophotometric standards \citep{Vernet2010, Hamuy1994}, mirror flexure
compensation, initial sky-subtraction and lastly the rectification and merging
of the orders. For the initial sky-subtraction, the background has been
estimated in regions adjacent to the object trace clear of contaminating
sources. Because of the broken ADC, for some objects there is a lot of curvature
in the object trace along the dispersion axis of the slit. This means that for
some bursts, the initial sky-estimate has been made from a limited number of
pixels in the spatial direction. By doing an initial subtraction of the sky
background on the un-rectified image we ensure that the bulk of the sky
background is not redistributed by the rectification process.

The image is rectified onto an equidistant grid with a dispersion sampling of
0.02 nm/pixel and a 0.16 "/pixel spatial sampling for the UVB and VIS arm and
0.06 nm/pixel with a 0.21 "/pixel in the NIR arm.  Because the tabulated
resolution is a lower limit, we choose a sampling of 0.02 nm/pixel to ensure
that the bluest part of neither of the arms have a sampling lower than the
Nyquist sampling rate of 2 pixels per resolution FWHM.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as a single observation and then subsequently combined to form a
single image. We employ this strategy so that we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images where the noise in each pixel is dominated by
Poisson noise it is important to estimate the background variance in a large enough
region, so that any correlation between the signal and the weights are removed.
To this end, the weight map is generated by a running median window over the
variance map, where the trace has been masked and width of the window is chosen
to be wide enough for median variance to be generated on the basis of several
hundred pixels. This weighting scheme automatically also optimally combines
images of different exposure times or images where the background is varying,
which is often the case when a burst has been observed close to twilight

@@

Since X-shooter is installed at the VLT Cassegrain focus it is prone to
flexures. The flexures modify the projection of the slit on the detector with
respect to the one obtained in daytime calibration. This requires a modification
of the wavelength solution in order to correctly process the night-time data.
Part of this correction is performed by the pipeline using the frames taken
during the X-shooter Active Flexure Compensation procedure \footnote{X-shooter
	User Manual available at
	https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}. The
remaining offset is corrected by cross-correlating the observed sky spectrum
with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum,
estimated as the mode of all flux values, has been subtracted. To get the
correct seeing PSF with which to convolve the synthetic sky an initial
refinement of the wavelength solution has been obtained by cross-correlating
the observed sky with an unconvolved synthetic sky. This preliminary wavelength
calibration is applied to the observed sky. The synthetic spectrum is then
convolved with an increasing seeing PSF and the width that minimizes $\chi^2$
with the updated observed sky is chosen to be the effective sky-PSF. Using the
synthetic sky with the matched resolution, a final wavelength calibration can
then be calculated by cross-correlating the observed sky with the correctly
broadened sky spectrum, as a function of a velocity offset. Both a
multiplicative and an additive offset to the wavelength calibration has been
tested, but in terms of $\chi^2$, the model with only a multiplicative offset is
preferred. The resulting offsets, which were smaller than 0.01 nm in the UVB and
VIS data and smaller than 0.05 nm in the NIR spectra, but changing over short
period of time were applied to the corresponding spectra. Using the convolved
synthetic sky, the pixels containing the brightest sky lines have been added to
the bad pixel map.

@@

To extract the afterglow spectrum from the rectified 2D-image, several
techniques have been employed based on the brightness of the afterglow and the
complexity of the objects entering the slit. Due to the malfunctioning ADC, see
Sect. \ref{obs}, the spectral trace changes position across the slit in the
spatial direction as a function of wavelength. For a large fraction of the
observed bursts, using a single aperture for the spectral extraction is
inadequate due to the large amount of background that would then enter the slit.
To optimally select the extraction regions we therefore need to model the trace
position.

To get the shape and the position of the spectral PSF as a function of location
on the image, we need to chose a model which can represent how the light falls
on the slit. We know from \citet{Trujillo2001} that the Moffat function
\citep{Moffat1969} adequately describes an imaging PSF due to atmospheric
turbulence, but because of aberrations in the dispersion elements and the
rectification process, the PSF we are trying to model is different from this
profile. To allow for flexibility in the model, we have chosen the Voigt
function as a model for the spectral PSF and we describe how this is evaluated
in App. \ref{voigt}. Since additionally, the host galaxy could also have a
contribution to the image profile, this choice allows for the required freedom if
additional flux is in the wings of the profile.

@@

Based on the signal-to-noise of the afterglow continuum, the 2D-image has been
binned down in the spectral direction to a number of elements that allows for an
accurate tracing of the PSF, typically 200 bins for moderate signal-to-noise.
For each of the bins, using the analytically guided guess position, the spectral
PSF has been fit using the unweighted chi-squared minimization algorithm
implemented in \texttt{scipy.optimize.curve\_fit} \citep{scipy}. Since we know
that the trace varies slowly as a function of wavelength, we have then fitted a
low-order polynomial to the fit parameters as a function of wavelength, which
allows us to evaluate the spectral PSF at all wavelengths and in this way
accurately model the entire spectral PSF.

Equipped with a model for how the light is distributed across the entire
dispersion direction, we can  employ the optimal extraction algorithm
\citep{Horne1986}, which weights the extraction aperture by the spectral profile,
or alternatively sum all pixels within 1 FWHM of the modeled profile. Where
possible, we have used the optimal extraction. In cases where the trace is very
weak, even in the binned images, an aperture has been selected manually which
covers emission lines, if present, and when nothing is immediately visible, the
entire nodding window. The error- and bad pixel maps are in all cases propagated
throughout the extraction.

@@

The spectra are corrected for Galactic extinction using the $E(B-V)$ value from
the dust maps of \citet{Schlegel1998} with the update in
\citet{Schlafly2011}\footnote{Queried from
	http://irsa.ipac.caltech.edu/applications/DUST/index.html using
	\citet{astroquery}}, and the extinction curve by \cite{Cardelli1989} with a
total to selective extinction $R_V = 3.1$. The wavelengths of the extracted
1D-spectra are moved to vacuum, corrected for barycentric motion, and the
wavelength recalibration described in Sect. \ref{wavecal} is applied. Pixels
with pixel-to-pixel variation larger than $50 \sigma$ are additionally added to
the bad pixel map.

@@

For all earth-based telescopes, the light first has to pass through Earths
atmosphere, where the atmospheric content and conditions make an imprint on the
received spectrum. These telluric features can be corrected for in a multitude
of ways. We employ a prioritized list of methods here, depending on the
availability of the different methods. Since the observation are often taken at
odd times under varying conditions, this prioritized list ensures that we are
always doing the optimal correction.

@@

In the last case, where the object is neither bright enough, or there for some
reason have not been observed a telluric standard, we rely on the synthetic sky
model by \citet{Noll2012} and \citet{Jones2013} for which we generate a
synthetic transmission spectrum,  where the ambient parameters for the
observations have been used.

@@

The afterglow spectra described in this paper are obtained in
Target-of-Opportunity (override) mode. In most cases there are therefore little
possibility to tweak slit widths to the seeing at the time of observations (i.e.
to optimize spectral resolution and signal to noise), and almost all our data is
therefore taken with a fixed set of slit widths and binning, described above. In
a fair number of cases, the seeing full width at half maximum (FWHM) is
considerably smaller than the silt width, and the delivered spectral resolution
will then be determined by the seeing rather than slit width, as afterglows are
point sources (this is evidently not the case for extended sources, e.g for host
galaxies). The delivered resolution for slit-width-dominated spectra
post-reduction and extraction can easily be determined from the bright sky
emission lines. For afterglow spectra with very high signal to noise, the
delivered spectral resolution can at times be determined from the science data
themselves. However, in the presence of multiple velocity components in
absorption, other forms of line broadening, and a lack of lines at some
redshifts, this is difficult to do at poorer signal to noise ratios (the
majority of spectra in our sample). A broad starting value for the expected
resolution will help fitting of these spectra, and can be important in upper
limit determination, and for this reason we construct a crude relation between
the seeing and the delivered resolution at our slit width, binning, and
reduction pipeline settings.

@@

All the spectra are made available at
\url{http://www.dark-cosmology.dk/~jselsing/XSGRB} and additionally through the
ESO archive in the form of phase 3 material
\footnote{\url{http://archive.eso.org/wdb/wdb/adp/phase3_main/form}}. This
release includes for completeness the spectra of GRBs outside the statistical
sample and includes both prompt afterglow observations as well as late time
observations of the associated hosts and represents \textit{all} afterglow
spectra of GRBs carried out by the X-shooter spectrograph since the
commissioning of the instrument, \startdate, and until the end of the last
period of the program 098.A-0055, \termdate~and thus constitutes 8 years of GRB
afterglow observations with X-shooter. An overview of all the spectra provided
here are given in Table. \ref{tab:sample_overview}. For each burst, each
individual observation is provided in a separate reduction and in cases where
observations have been repeated for an increased S/N or to follow the temporal
evolution, a combined spectrum is also provided. No attempt has been made to
join the spectroscopic arms, so for each observation, three spectra are provided
in separate files.

@@

Each file contains 7 columns with following contents and descriptions:
\begin{itemize}
	\item WAVE - Observed wavelength in vacuum, corrected for barycentric motion and drifts in the wavelength solution ($\mathrm{\AA}$).
	\item FLUX - Observed flux density ($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-2} \mathrm{\AA}^{-1}$)
	\item ERR - Associated flux density error ($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-2} \mathrm{\AA}^{-1}$)
	\item QUAL - Bad pixel map, where a value different from zero indicates a bad pixel.
	\item CONTINUUM - Continuum estimate based on Sect. \ref{continuum} ($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-2} \mathrm{\AA}^{-1}$)
	\item CONTINUUM\_ERR - Relative error on the continuum estimate 
	\item TELL\_CORR - Inverse transmission spectrum. Multiply FLUX and ERR column with this column to correct for telluric absorption. 
\end{itemize}

@@

In this section, we describe the efficiency of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
obtained sample is representative for the full \textit{Swift} sample. An
important note is that here we include \textit{all} GRBs after \startdate, that
have been observed with X-shooter, while only a subset of these constitutes our
\textit{statistical} sample. The statistical sample is based on the selection
criteria described in Sect. \ref{samplecrit}. Some bursts not fulfilling the
sample criteria have been followed up due to interesting characteristics, e.g.
curiosity properties of their light curves, their brightness, etc. These bursts
are not discussed as part of the investigation of the statistical properties of
the GRB population. A prime example of a burst outside the statistical sample is the
bright \textit{INTEGRAL} burst GRB~161023A, that turned to be a rich spectrum with at
least 11 intervening absorption systems.

@@

This spectrum is an attempt at a late-time host redshift measurement for
GRB~141031A. The spectrum is taken long after the burst has faded and consists
of $4\times600$ s spectroscopic integration. In the 2D spectrum are two sources,
which are both offset from the attempted host and are thus likely foreground
objects. The spectrum does not contain anything that can be used to measure a
redshift from.
